Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-05 paper/paper_draft.qmd
Updates to methods section

The spatial extent of reefs represented in this investigation follows reef definitions
available via the Great Barrier Reef Marine Park Authority (GBRMPA) [@GBRMPA2022]. The reef
definitions define 3806 reefs across the GBR. An indicative map of the GBR marine park
including the Marine Park Management areas [@GBRMPA2007] and the corresponding bioregion for
each reef is shown in [@fig-region-map] [@GBRMPA2024]. Estimates of coral carrying capacity
for each reef were taken from prior analyses undertaken in Bozec et al. [-@bozec2022].

Depth characteristics for each reef were extracted using 10m resolution bathymetry data made
available in GBRMPA [-@GBRMPA2020]. While the bathymetry data adopted for the given
assessment are said to be accurate up to 15m from Mean Sea Level [@GBRMPA2020], there are
many sources of uncertainty in the depth estimates used for this study. Notably, the adopted
depth estimates were derived from bathymetries based on satellite imagery. The bathymetry
data can be limited in their alignment with the established GBRMPA reef outlines used to
characterise reef locations and their areal extent. For each reef, bathymetry data were
spatially matched with the 3806 reef outlines through geometric intersections.

A Quality Assessment and Control process identified reefs with potential data limitations:
reefs with no overlapping bathymetry data, reefs with minimum depth values above sea level
(indicating potential data artifacts), and reefs where bathymetry pixels covered less than
5% of the reef polygon area. This process removed 842 reefs, or 22%, from the initial set of
3806 reefs. Furthermore, reefs that are not associated with GBRMPA Management areas
[@GBRMPA2007] or reefal bioregions [@GBRMPA2024] were also removed from the investigation.
Bioregions which contained fewer than three reefs after these processes were also excluded.
As a result, 1377 reefs were excluded (36%), leaving a total of 2429 remaining. All depth
values were converted to positive values (depths below sea level). The median depth was used
to provide an indicative depth value for each reef. Further limitations are outlined in the
Discussion. Code to extract depth estimates, and the resulting depth estimates themselves,
are available at [@iwanaga2025]. To ensure effective data analysis, reefs with incoming or
outgoing connectivity values below the 5th percentile were removed from post-hoc analyses,
leaving 2197 reefs remaining for display in results.

![Map of the Great Barrier Reef region, including GBRMPA Management areas (polygons) and all
considered reefs (points). Colour-coded polygons outline GBR marine park management regions,
where orange represents the Far Northern management region, green represents
Cairns/Cooktown, red covers Townsville/Whitsunday and blue covers Mackay/Capricorn
[@GBRMPA2007]. Reef centroids are marked individually for visual clarity and are coloured
according to its bioregion [@GBRMPA2024]. All 2197 reefs included in this investigation are
represented in this figure. Annotations represent major coastal settlements along the
GBR.](../figs/region_map.png){#fig-region-map fig-align="center" width="14.8cm"}

@@

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled
1km^2^ resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GFDL-CM4. The selected GCMs are a part of the
Sixth Phase of the Coupled Model Intercomparison Project (CMIP6), spanning four 'model
development families' as defined by Brunner et al., [-@brunner2020].

These GCMs were selected based on evaluation by Grose et al., [-@grose2023], indicating
their ability to recreate historical atmospheric temperature and precipitation conditions in
the Australasian region. They span a range of Equilibrium Climate Sensitivity (ECS) values
within the Intergovernmental Panel on Climate Change (IPCC) assessed likely range. The IPCC
defines ECS as the long-term warming caused by a doubling of carbon dioxide above its
pre-industrial concentration [@forster2021]. While this assessment does not consider oceanic
data, Grose et al. [-@grose2023] provide a region-specific selection mechanism which we use
as an indicator of general performance of GCMs for the region. An overview of the selected
GCMs is provided in @tbl-gcms. Note here that Grose et al., [-@grose2023] assessed
GFDL-ESM4, whereas Dixon et al., [-@dixon2022; -@dixon2023] provides data based on GFDL-CM4.
This substitution is assumed appropriate given they share the same fundamental
representation of the physical climate system.

Shared Socio-economic Pathway (SSP) 2-4.5 was used to represent the plausible emission
pathway. It represents a moderate emissions scenario with continued CO~2~ emissions
comparable to present day until 2050, after which emissions gradually decline
[@hausfather2020; IPCC -@climate2023]. The selected GCMs represent 2.5°C to 4.66°C of global
atmospheric warming in the 2081-2100 period relative to 1850-1900. The use of these GCMs
therefore allow for a range of possible climate futures to be explored, capturing
uncertainty in future responses to CO~2~ increases (@sfig-ecs; @fig-methods-dhws;
[@grose2023; @shepherd2018]) and enables the assessment of modelled coral reef behaviour
over a broad range of independently derived climate futures [@abramowitz2019; @brunner2020].

Daily SST data [@dixon2023] were used to calculate climatological baselines and “HotSpot
Anomaly” values for each pixel following NOAA methodology [@liu2017; @skirving2020].
Additional information about DHW calculation methods is provided in the supplemental
material. DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the
12-week rolling sum of HotSpot SST Anomaly values that are greater than 1°C, expressed in
units of °C-weeks. The 12-week accumulation period reflects the cumulative nature of coral
thermal stress. Brief temperature spikes may be tolerable, but sustained or repeated
exposure to elevated temperatures causes progressively greater physiological stress and
bleaching severity. DHW values integrate both the magnitude and duration of thermal stress:
a moderate anomaly (e.g., 1°C) sustained for 12 weeks produces the same DHW value as a
severe anomaly (e.g., 3°C) lasting 4 weeks, though the biological impacts may differ. DHW
estimates for each indicative reef (see @GBRMPA2022) were then determined by taking the mean
DHW value for all pixels touching each reef outline. The process described here was
conducted for all 2197 reefs to create DHW projections spanning from 2025 to 2075.

@@

Information for each GCM used in this investigation including Equilibrium Climate
Sensitivity (ECS) value, interpreted relative severity and where results for an individual
GCM results are found. GCMs in context with assessed plausible ranges of ECS values are
displayed in [@sfig-ecs]. ACCESS-ESM1-5 has an ECS closest to the midpoint of the
very-likely ECS range and was taken as an intermediate model with results displayed in the
main text.
:::

![Degree Heating Week time series (in °C - Weeks) for the GCMs selected. Colours represent
the different Management areas across the Great Barrier Reef. Green represents
Cairns-Cooktown, orange represents Far-Northern, blue represents Mackay-Capricorn and red
represents the Townsville-Whitsunday Management area. Solid lines represent the
management-area median time series and bands represent the 95th percentile confidence
intervals.](../figs/methods_dhw_timeseries.png){#fig-methods-dhws fig-align="center"
width="14.82cm"}

@@

An exploratory scenario modelling approach was adopted for this investigation. The method
involves exploring a wide range of potential futures and model parameterisations under
different assumptions of environmental (climatic) and ecological conditions. Exploratory
modelling approaches have gained favour in recent years to explore the implications of deep
uncertainty; where there is little or no agreement on the key drivers of a system, the
extent of the relationship, how best to represent them or their probability distributions
[@conradt2024; @lempert2013]. The approach embraces uncertainty as an inherent
characteristic of complex environmental systems [@lempert2010; @lempert2013] in an attempt
to address limitations of calibrated ecosystem models with a singular set of parameter
values [@lubianabotelho2025]. Exploratory scenario modelling is a core methodology
underpinning the development of robust and adaptive management strategies [@haasnoot2013].

Rather than attempting to forecast precise outcomes, the exploratory approach aims to
systematically search many possible futures to identify patterns that lead to outcomes of
interest. The approach allows for the identification of robust management strategies that
perform well across a diversity of future conditions rather than attempting to identify a
single most optimal strategy. Concurrently, assessing the wide swathe of scenarios allows
contextual understanding of the conditions in which strategies may fail. Multiple
interacting and compounding sources of uncertainty, and a lack of broad agreement as to
their likelihood, create inherently uncertain outcomes. Here, exploratory modelling was
applied to identify the environmental and reef characteristics, such as thermal stress,
depth, and connectivity, that support favourable outcomes in reef recovery and resilience.

@@

Larval transition probability matrices were generated using the EReefs Great Barrier Reef
1km resolution hydrodynamic model and particle tracking simulations. Connectivity data from
24 days in total were used in this investigation to estimate coral larval flows. These data
span six years from 2015 to 2020. The years 2015, 2016, 2018 and 2019 are each represented
by connectivity matrices covering spawning periods over three days in each year (12 days in
total), while 2017 and 2020 are each represented by six spawning days each, coming to a
total of 24 days of connectivity matrices. The values in these matrices indicate the
probability of larval transfer between each pair of reefs in the GBR, herein referred to as
connectivity strengths. Indicative connectivity strengths are implemented in ADRIA-CoralBlox
as the mean of all available data for each location (i.e., mean of all 24 spawning days).
While the connectivity strengths do not vary temporally within each ADRIA-CoralBlox
simulation, they do vary between reefs. This spatial variability enables investigations into
the influence of connectivity strength and its role in driving recovery processes.

To assess the impact of reef connectivity on modelled outcomes we summarise incoming
connectivity (weighted incoming connectivity), outgoing connectivity (outgoing connection
strength), the ratio of source-to-sink connectivity, and larval retention probability for
each reef. Weighted incoming connectivity is calculated by summing the product of each
connection strength incoming to a target reef (excluding retained larvae from the target
reef) and the mean timeseries absolute coral cover (km^2^) of each matching source reef.
This metric provides an approximation of overall larval flow into a reef, with high values
indicating the target reef is supplied high volumes of larvae via strong connections and low
values indicating the target reef has weak (or no) incoming connections or is supplied by
small source populations. A limitation of weighted incoming connectivity is that it does not
incorporate the proportion of the source reef's population that is reproducing and producing
larvae, instead taking the total coral cover area as an indicator of larval productivity.
Outgoing connection strength is calculated as the sum of all outgoing transition
probabilities from a target reef. Outgoing connection strength has a theoretical maximum of
one (all larvae reach settlement on sink reefs), however the large size of the GBR combined
with larval mortality considerations result in outgoing connection strength values much
lower than one. The ratio of source-to-sink connectivity ($O:I$) for a target reef
represents the strength of the reef's outgoing larval flow relative to it's incoming larval
flow. This is calculated as: $O:I = \frac{S_O/N_O}{S_I/N_I}$. Where $S$ represents the sum
of connectivity probabilities, $N$ represents the number of connections, $O$ denotes
outgoing connectivity from the target reef and $I$ denotes incoming connectivity to the
target reef. If a reef has no incoming connections, the $O:I$ equals the $S_O/N_O$ value.
Incoming connections in this calculation, do not directly represent the larval supply gained
as the breeding populations from which connections are coming from are not represented. A
source-to-sink ratio greater than 1 (a value greater than 0 on a log scale) indicates a reef
has greater outgoing than incoming larval flow. Reef larval retention probability indicates
the proportion of spawned larvae from a target reef that remain to settlement.

@@

Two separate systems were employed to explore scenarios: ADRIA (the Adaptive Dynamic Reef
Intervention Algorithms platform) and CoralBlox. ADRIA is a multi-criteria decision support
platform designed for the assessment and analysis of coral reef intervention strategies
under uncertainty [@iwanaga2023]. The ADRIA platform provides functionality for scenario
generation and analysis as well as providing the technical infrastructure to couple
environmental and ecosystem representations. CoralBlox is a computationally efficient coral
population model. It simulates functional/morphological groups of corals over their life
cycles, competing for available space over time [@ribeirodealmeida2024]. The coupled
framework is referred to as ADRIA-CoralBlox.

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of important note is that the representation of
bleaching mortalities followed Bozec et al., [-@bozec2022] albeit with a scaling coefficient
to incorporate the amelioration effect that depth may have, following Baird et al.,
[@baird2018]. This depth-scaling reduces the proportion of coral colonies impacted by high
heat stress as depth increases. The impacts of tropical cyclones, wave activity, Crown of
Thorns Starfish or coral disease were not included in this investigation, in order to focus
on assessing reef resilience under possible future warming scenarios.

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For the purpose of scenario assessment, the resulting coral
cover time series for each model run were scaled by the total area of each reef to represent
cover as a percentage of the total estimated reef area. The scenario-median time series for
each reef was used for further analyses. These time series of coral cover were then
clustered based on their temporal behaviour. Assessment metrics used in analyses can be
found in @stbl-metrics.

@@

The Complexity-Invariant Distance (CID) between each time series was calculated to identify
similarities in temporal behaviour [@batista2014]. The CID applies a correction factor to
the Euclidean distance between pairs of time series to obtain a measure of their similarity.
The adopted approach ensures reefs are clustered based on their temporal dynamics as opposed
to, for example, naively clustering by their mean coral cover. Time series clustering is
achieved through the use of k-medoids clustering [cf. @steinmann2020]. Further details on
the time series clustering approach are found in the supplementary materials.

Time series clustering was applied over the time frame between 2025 and 2075. The clustering
was conducted for three spatial scales: (1) the whole of GBR; (2) each GBR management area;
and (3) for each reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). The analyses were
conducted at the different spatial scales as, conceptually, the temporal behaviour
constituting "low", "medium" or "high" coral cover may differ at the different spatial
scales. As it was, analyses at the whole-of-GBR scale were found to lead to identical
conclusions regarding reef characteristics which influence coral cover and so for brevity
are not shown here.

@@

Partial Dependence Analysis was conducted to understand the relationship between features
and bioregion-scale time series cluster predictions using a Random Forest (RF) classifier
[@friedman2001]. The analysis utilised the reef characteristics of interest as the
predictors, including the estimated median reef depth, mean DHW for the given GCM,
log10-weighted incoming connectivity, outgoing connectivity, and log10-transformed
outgoing-to-incoming connectivity ratio. The full dataset consisted of the time series
cluster assignment for each of the 2197 reefs under the five GCMs explored, for a total of
10,985 samples. The dataset was randomly split into training (60%) and testing (40%) subsets
to train and assess classification performance. The RF classifier achieved 80% accuracy on
the training set and 60% on the test set.

A 3-fold cross-validation was also conducted, arriving at a similar 59% accuracy rate. F1
performance scores for Low, Medium, and High clusters were determined to be 0.70, 0.43 and
0.64 respectively, with precision of 0.62, 0.51 and 0.66 when evaluated with the test set
(consisting of 4394 samples) (@stbl-confusion-matrix; @stbl-rf-cluster-performance). Lower
performance with the Medium cluster reflects the inherent difficulty of classifying
transitional states in natural systems. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier, while not
accurate, was deemed sufficiently informative for the purpose of assessing the influence of
reef characteristics.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically varied (one-at-a-time) across a grid of
50 equally spaced points spanning the predictor's range, while all other features were left
at their modelled values. The procedure creates (1) a modified test dataset, where the
target predictor was set to the grid value for all samples, (2) probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class across all test observations. Use of the median predicted
probability was also assessed but was found not to substantially influence the
interpretation of results (and so are not shown here). The process isolates the marginal
effect of each predictor by averaging over the natural distribution of all other
representative reef characteristics in the test dataset. Additionally, to assess possible
interaction between reef depth and connectivity metrics, two-way partial dependence analysis
was performed on grids along the ranges of depth and weighted incoming connectivity, and
depth and outgoing connection strength.

@@

Using coral cover time series for each reef we calculated the number of years each reef
maintains coral cover above a given percent threshold. The threshold represents the level of
live coral cover necessary to maintain a positive carbonate budget (denoted here as
$\theta_{0}$ to indicate the budget equilibrium). Greater number of years above $\theta_{0}$
suggests greater resilience and opportunity for recovery.

Recent analyses suggest a general threshold of \~17% live coral cover in the case of the
mid-shelf and offshore GBR reefs [@desbiens2021]. Given in situ carbonate budgets are
influenced by (1) a range of processes; (2) the variety of approaches that could be used to
measure those processes; and (3) the uncertainty surrounding these, it would be unreasonable
to assume all reefs adhere to a single precise $\theta_{0}$ value. Concurrently, it is
difficult to ascertain a suitable $\theta_{0}$ value for each individual reef.

@@

The relationships between the number of years above $\theta_{0}$ and characteristics such as
median reef depth and log10 total connectivity strength were then assessed with Spearman's
rank correlation. Spearman's correlation coefficient was adopted as reef depth and
connectivity strength distributions are expected to be non-normal and highly skewed.
Furthermore, reefs are naturally of different types and experience different conditions, and
so the monotonic trends (whether the relationship consistently increase or decrease) rather
than the exact quantitative relationship was of interest for this analysis.

>>>>

The spatial extent of reefs represented in this investigation follows reef definitions
available via the Great Barrier Reef Marine Park Authority (GBRMPA), defining 3806 reefs
across the GBR [@GBRMPA2022] (@fig-region-map). Estimates of coral carrying capacity for
each reef were taken from prior analyses [@bozec2022].

Depths for each reef were extracted using 10m resolution bathymetry data made available in
GBRMPA [-@GBRMPA2020]. While the bathymetry data are said to be accurate up to 15m from Mean
Sea Level [@GBRMPA2020], there are many sources of uncertainty in the depth estimates used
for this study. Notably, bathymetry data were derived from satellite imagery. The bathymetry
data can be limited in their alignment with the established GBRMPA reef outlines used to
characterise reef locations and their areal extent. For each reef, bathymetry data were
spatially matched with the 3806 reef outlines through geometric intersections. All depth
values were converted to positive values (depths below sea level). The median depth was used
to provide an indicative depth value for each reef. Further limitations are outlined in the
Discussion. Code to extract depth estimates, and the resulting depth estimates themselves,
are available at [@iwanaga2025].

A Quality Assessment and Control process identified reefs with potential data limitations:
reefs with no overlapping bathymetry data, reefs with minimum depth values above sea level
(indicating potential data artifacts), and reefs where bathymetry pixels covered less than
5% of the reef polygon area. This process removed 842 reefs, or 22%, from the initial set of
3806 reefs. Furthermore, reefs that are not associated with GBRMPA Management areas
[@GBRMPA2007] or reefal bioregions [@GBRMPA2024] were removed. Bioregions which contained
fewer than three reefs after these processes were also excluded. As a result, 1377 reefs
were excluded, leaving a total of 2429 remaining for time series clustering. To ensure
effective data analysis, reefs with incoming or outgoing connectivity values below the 5th
percentile were removed from post-hoc analyses, leaving 2197 reefs remaining for display in
results (@fig-region-map).

![Map of the Great Barrier Reef region, including GBRMPA Management areas (polygons) and all
considered reefs (points). Colour-coded polygons outline GBR marine park management regions:
Far Northern (orange), Cairns-Cooktown (green), Townsville-Whitsunday (blue) and
Mackay-Capricorn (red) [@GBRMPA2007]. Reef centroids are marked individually for visual
clarity and are coloured by bioregions [@GBRMPA2024]. All 2197 reefs included in this
investigation are represented in this figure. Annotations represent major coastal
settlements along the GBR.](../figs/region_map.png){#fig-region-map fig-align="center"
width="14.8cm"}

@@

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled
1km^2^ resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GFDL-CM4. The selected GCMs are a part of the
Sixth Phase of the Coupled Model Intercomparison Project (CMIP6), spanning four 'model
development families' as defined by Brunner et al., [-@brunner2020]. Shared Socio-economic
Pathway (SSP) 2-4.5 was used to represent the plausible emission pathway. It represents a
moderate emissions scenario with continued CO~2~ emissions comparable to present day until
2050, after which emissions gradually decline [@hausfather2020; IPCC -@climate2023].

These GCMs were selected based on evaluation by Grose et al., [-@grose2023], indicating
their ability to recreate historical atmospheric temperature and precipitation conditions in
the Australasian region. They span a range of Equilibrium Climate Sensitivity (ECS) values
within the Intergovernmental Panel on Climate Change (IPCC) assessed likely range. The IPCC
defines ECS as the long-term warming caused by a doubling of carbon dioxide above its
pre-industrial concentration [@forster2021]. While Grose et al. [-@grose2023] do not
consider oceanic data, they provide a region-specific selection mechanism which is used as
an indicator of general performance of GCMs for the region. An overview of the selected GCMs
is provided in @tbl-gcms. Note here that Grose et al., [-@grose2023] assessed GFDL-ESM4,
whereas Dixon et al., [-@dixon2022; -@dixon2023] provides data based on GFDL-CM4. This
substitution is assumed appropriate given they share the same fundamental representation of
the physical climate system. The use of these GCMs therefore allows for a range of possible
futures to be explored, capturing uncertainty in future responses to CO~2~ increases
(@fig-methods-dhws; @sfig-ecs;[@grose2023; @shepherd2018]) and enables the assessment of
modelled coral reef behaviour over a broad range of independently derived climate futures
[@abramowitz2019; @brunner2020].

Daily SST data [@dixon2023] were used to calculate climatological baselines and “HotSpot
Anomaly” values for each pixel following NOAA methodology [@liu2017; @skirving2020].
Additional information about DHW calculation methods is provided in the supplemental
material. DHW values on a per-pixel basis at 1km^2^ resolution were calculated as the
12-week rolling sum of HotSpot SST Anomaly values that are greater than 1°C, expressed in
units of °C-weeks. The 12-week accumulation period reflects the cumulative nature of coral
thermal stress. Brief temperature spikes may be tolerable, but sustained or repeated
exposure to elevated temperatures causes progressively greater physiological stress and
bleaching severity. DHW values integrate both the magnitude and duration of thermal stress:
a moderate anomaly (e.g., 1°C) sustained for 12 weeks produces the same DHW value as a
severe anomaly (e.g., 3°C) lasting 4 weeks, though the biological impacts may differ. DHW
estimates for each indicative reef (see @GBRMPA2022) were then determined by taking the mean
DHW value for all pixels touching each reef outline. The process described here was
conducted for all 2197 reefs to create DHW projections spanning from 2025 to 2075.

@@

Information for each GCM used in this investigation including Equilibrium Climate
Sensitivity (ECS) value, interpreted relative severity and where results for an individual
GCM results are found. GCMs in context with assessed plausible ranges of ECS values are
displayed in [@sfig-ecs]. ACCESS-ESM1-5 has an ECS closest to the midpoint of the likely ECS
range and was taken as an intermediate model with results displayed in the main text.
:::

![Degree Heating Week time series (in °C - Weeks) for the GCMs selected. Colours represent
the different Management areas across the Great Barrier Reef: Far Northern (orange),
Cairns-Cooktown (green), Townsville-Whitsunday (blue) and Mackay-Capricorn (red). Solid
lines represent management-area median time series and bands represent 95th percentile
confidence intervals.](../figs/methods_dhw_timeseries.png){#fig-methods-dhws
fig-align="center" width="14.82cm"}

@@

An exploratory scenario modelling approach was adopted for this investigation. The method
involves exploring a wide range of potential futures and model parameterisations under
different assumptions of environmental (climatic) and ecological conditions. Exploratory
modelling approaches have gained favour in recent years to explore the implications of deep
uncertainty; where there is little or no agreement on the key drivers of a system, the
extent of the relationship, how best to represent them or their probability distributions
[@conradt2024; @lempert2013]. The approach embraces uncertainty as an inherent
characteristic of complex environmental systems [@lempert2010; @lempert2013] in an attempt
to address limitations of calibrated ecosystem models that use a singular set of parameter
values [@lubianabotelho2025]. Exploratory scenario modelling is a core methodology
underpinning the development of robust and adaptive management strategies [@haasnoot2013].

@@

The coral larval connectivity model used in this study comprises 3D hourly velocities from
the 1 km grid resolution hydrodynamic model (GBR-1 version 2.0) [@steven2019], as input to
OceanParcels [@delandmeter2019], a particle tracking model. The connectivity model was used
to simulate larval dispersal and settlement of broadcast-spawning Acropora corals in the
GBR, following annual mass spawning events in 2015, 2016, 2018, 2019 and split-spawning
events in 2017 and 2020. Split spawning occurs when the annual reproductive event of a coral
population is split over two consecutive months. Coral mass spawning usually occurs 4-6 days
after the full moon during the spring-summer transition in the GBR with variations across
species and locations. The spawning days identified are: 2015 - November 30, December 1 and
2; 2016 - November 18, 19 and 20; 2017 - November 8, 9 and 10, December 8, 9 and 10; 2018 –
November 18, 19, 20; 2019 – November 16, 17, 18; 2020 – November 5, 6 and 7, December 4, 5,
6.

For each spawning night, particles were released from random locations within each reef
polygon as neutrally buoyant virtual larvae every 27 seconds from 8:00 pm to 11:00 pm at
2.25 m below the surface. In each of the reef polygons, 401 particles were released
resulting in a total of 1,526,206 particles every night of spawning. The model was run for
30 days with a 3-minute timestep, and the locations of the released particles were updated
every 15 minutes until the end of the simulation. Larvae were assumed to become competent to
settle from 4 to 28 days post release with a daily mortality rate of 0.4 [@ani2024], which
was implemented in 15-minute timestep as probabilistic mortality, corresponding to 0.0053
mortality rate per timestep. To incorporate the stochastic sensory behaviour of larvae
during settlement, a 0.01 probability of settlement is assumed when larvae are over a reef
during the competency period per 15-minute timestep. The information obtained on larvae
settlement was used to compute a connectivity matrix among reef polygons for each spawning
night (see Ani et al. [-@ani2024] for more details). A total of 24 connectivity matrices
were generated for the spawning nights that were considered.

Larval transition matrices indicate the probability of larval transfer between pairs of
reefs in the GBR, herein referred to as connectivity strengths. Indicative connectivity
strengths are implemented in ADRIA-CoralBlox as the mean of matrices for each location
(i.e., mean of all 24 spawning days). While the connectivity strengths do not vary
temporally within each ADRIA-CoralBlox simulation, spatial variability enables
investigations into the influence of connectivity strength and its role in driving recovery
processes.

To assess the impact of reef connectivity on modelled outcomes, connectivity was summarised
by incoming connectivity (weighted incoming connectivity), outgoing connectivity (outgoing
connection strength), the ratio of source-to-sink connectivity, and larval retention
probability for each reef. Weighted incoming connectivity was calculated by summing the
product of each connection strength incoming to a target reef (excluding retained larvae
from the target reef) and the mean timeseries absolute coral cover (km^2^) of each matching
source reef. This metric provides an approximation of overall larval flow into a reef, with
high values indicating the target reef is supplied high volumes of larvae via strong
connections and low values indicating the target reef has weak (or no) incoming connections
or is supplied by small source populations. A limitation of weighted incoming connectivity
is that it does not incorporate the proportion of the source reef's population that is
reproducing and producing larvae, instead it takes total coral cover area as an indicator of
larval productivity. Outgoing connection strength was calculated as the sum of all outgoing
transition probabilities from a target reef. Outgoing connection strength has a theoretical
maximum of one (all larvae reach settlement on sink reefs), however the large size of the
GBR combined with larval mortality result in much lower outgoing connection strength values.
The ratio of source-to-sink connectivity ($O:I$) for a target reef represents the strength
of the reef's outgoing larval flow relative to it's incoming larval flow. This is calculated
as: $O:I = \frac{S_O/N_O}{S_I/N_I}$. Where $S$ represents the sum of connectivity
probabilities, $N$ represents the number of connections, $O$ denotes outgoing connectivity
from the target reef and $I$ denotes incoming connectivity to the target reef. If a reef has
no incoming connections, the $O:I$ equals the $S_O/N_O$ value. Incoming connections in this
calculation, do not directly represent the larval supply gained as the breeding populations
from which connections are coming from are not represented. A source-to-sink ratio greater
than 1 (a value greater than 0 on a log scale) indicates a reef has greater outgoing than
incoming larval flow. Reef larval retention probability indicates the proportion of spawned
larvae that settle on their natal reefs.

@@

Two systems were employed to explore scenarios: ADRIA (the Adaptive Dynamic Reef
Intervention Algorithms platform) and CoralBlox, referred to as the coupled framework
ADRIA-CoralBlox. ADRIA is a multi-criteria decision support platform designed for the
assessment and analysis of coral reef intervention strategies under uncertainty
[@iwanaga2023]. The ADRIA platform provides functionality for scenario generation and
analysis as well as providing technical infrastructure to couple environmental and ecosystem
representations. CoralBlox is a computationally efficient coral population model. It
simulates functional/morphological groups of corals over their life cycles, competing for
available space over time [@ribeirodealmeida2024].

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of note is that the representation of bleaching
mortalities followed Bozec et al., [-@bozec2022] albeit with a scaling coefficient to
incorporate the amelioration effect that depth may have, following Baird et al.,
[@baird2018]. This depth-scaling reduces the proportion of coral colonies impacted by high
heat stress as depth increases. The impacts of tropical cyclones, wave activity, Crown of
Thorns Starfish or coral disease were not included in this investigation in order to focus
on assessing reef resilience under possible future warming scenarios.

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For scenario assessment, coral cover time series for each
model run were scaled to represent cover as a percentage of the total estimated reef area.
The scenario-median time series for each reef was used for further analyses. Time series of
coral cover were then clustered based on their temporal behaviour. Assessment metrics used
in analyses can be found in @stbl-metrics.

@@

The Complexity-Invariant Distance (CID) between each time series was calculated to identify
similarities in temporal behaviour [@batista2014]. The CID applies a correction factor to
the Euclidean distance between pairs of time series to obtain a measure of their similarity.
The adopted approach ensures reefs are clustered based on their temporal dynamics as opposed
to, for example, naively clustering by their mean coral cover. Time series clustering is
achieved using k-medoids clustering [cf. @steinmann2020]. Further details on the time series
clustering approach are found in the supplementary materials.

Time series clustering was applied over the time frame between 2025 and 2075. The clustering
was conducted at three spatial scales: (1) the whole of GBR; (2) each GBR management area;
and (3) for each reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were
conducted at the different spatial scales as, conceptually, the temporal behaviour
constituting "low", "medium" or "high" coral cover may differ at different spatial scales.
As it was, analyses at the whole-of-GBR scale were found to lead to identical conclusions
regarding reef characteristics which influence coral cover and so for brevity are not shown
here.

@@

Partial Dependence Analysis (PDA) was conducted to understand relationships between features
and bioregion-scale time series cluster predictions using a Random Forest (RF) classifier
[@friedman2001]. The analysis utilised the reef characteristics of interest as the
predictors, including the estimated median reef depth, mean DHW for the given GCM,
log10-weighted incoming connectivity, outgoing connectivity, and log10-transformed
outgoing-to-incoming connectivity ratio. Conceptually, PDA is related to a Regional
Sensitivity Analysis [@hornberger1981; @pianosi2016], as model results are assessed to map
the influence of X - the predictors or model factors - on Y - the modelled outcomes
according to a metric of interest.

The full dataset consisted of the time series cluster assignment for each of the 2197 reefs
under the each GCM (totalling 10,985 samples). The dataset was randomly split into training
(60%) and testing (40%) subsets to train and assess classification performance. The RF
classifier achieved 80% accuracy on the training set and 60% on the test set.

A 3-fold cross-validation was also conducted, arriving at a similar 59% accuracy rate. F1
performance scores for Low, Medium, and High clusters were determined to be 0.70, 0.43 and
0.64 respectively, with precision of 0.62, 0.51 and 0.66 when evaluated with the test set
(consisting of 4394 samples) (@stbl-confusion-matrix; @stbl-rf-cluster-performance). Lower
performance with the Medium cluster reflects the inherent difficulty of classifying
transitional states in natural systems. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier, while not
accurate, is still sufficiently informative for the purpose of assessing the influence of
reef characteristics.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically varied (one-at-a-time) across a grid of
50 equally spaced points spanning the predictor's range, while all other features were left
at their modelled values. The procedure creates (1) a modified test dataset, where the
target predictor was set to the grid value for all samples, (2) probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class across all test observations. The process isolates the marginal
effect of each predictor by averaging over the natural distribution of all other
representative reef characteristics in the test dataset. Additionally, to assess possible
interaction between reef depth and connectivity metrics, two-way PDA was performed on grids
along the ranges of depth and weighted incoming connectivity, and depth and outgoing
connection strength.

@@

Using coral cover time series for each reef, the number of years each reef maintains coral
cover above a given percent threshold were calculated. The threshold represents the level of
live coral cover necessary to maintain a positive carbonate budget (denoted here as
$\theta_{0}$ to indicate the budget equilibrium). A greater number of years above
$\theta_{0}$ suggests greater resilience and opportunity for recovery.

Recent analyses suggest a general threshold of \~17% live coral cover in the case of the
mid-shelf and offshore GBR reefs [@desbiens2021]. Given in-situ carbonate budgets are
influenced by (1) a range of processes; (2) the variety of approaches that could be used to
measure those processes; and (3) the uncertainty surrounding these, it would be unreasonable
to assume all reefs adhere to a single precise $\theta_{0}$ value. Concurrently, it is
difficult to ascertain a suitable $\theta_{0}$ value for each individual reef.

@@

The relationships between the number of years above $\theta_{0}$ and characteristics such as
median reef depth and log10 total connectivity strength were assessed with Spearman's rank
correlation. Spearman's correlation was adopted as distributions of reef depths and number
of years above $\theta_{0}$ were expected to be non-normal and highly skewed. Furthermore,
reefs are naturally of different types and experience different conditions, and so the
monotonic trends (whether the relationship consistently increase or decrease) rather than
the exact quantitative relationship was of interest.
